Non-coding RNAs of bird genomes

### 2013-08-12 paper/manuscript.tex
Updated manuscript.

%% Talking points:
%% 1.     Although the majority of classical eukaryotic ncRNAs are conserved in birds, several notable classical ncRNAs are not. Homology searches in associated genes indicate these apparent "losses" are most likely due to divergence.
%% 2.     Of the 59 snoRNA conserved between humans and yeast, we find 45 are conserved in birds. Homology searches in associated genes for the remaining non-conserved snoRNAs indicate these apparent losses are indeed due to multiple losses in several bird lineages.
%% 3.     We find that 3 families of broadly conserved vertebrate microRNA families are lost in the avian lineage.  We also annotate a number of bird-specific microRNAs.

%% %% Draft Abstract
%% Here we present the results of a large-scale bioinformatic annotation of non-coding RNAs in 48 avian genomes. Our approach uses probabilistic models of hand-curated families from the Rfam database to infer conserved RNA families within each avian genome. We supplement these annotations with predictions from the tRNA annotation tool, tRNAscan-SE and microRNAs from miRBase. Although we show extensive conservation of classical ncRNAs (e.g., tRNAs and rRNAs) and more recently discovered ncRNAs (e.g., snoRNAs and miRNAs) in birds, we also demonstrate apparent ``losses'' in several RNA families. These include the divergence of some classical ncRNAs and the loss of several snoRNAs and microRNAs. These combined results illustrate the utility of applying homology based methods for annotating vertebrate genomes and illustrate many complex evolutionary patterns within the avian ncRNA cohort.

@@

%% 3.     Several well-characterised lncRNA families of known function are conserved between mammals and birds.
%% 4.     We discuss the conservation of HOX associated lncRNAs and several lncRNAs that have been implicated in processes associated with cancer.
%% In contrast, we show a signi_cant number of lncRNAs are surprisingly well conserved between birds and mammals including several intriguing cases where the reported mammalian lncRNA function is not conserved in birds.

@@

The 5.8S component of the ribosome in the turtle, turkey bustard,
hoatzin, flamingo, tropicbird, seriema, owl, cuckoo roller, trogon,
bee eater and falcon appears to have been lost (See
Figure~\ref{fig:1}). However, the genomes for these species are also
low-coverage. Therefore gaps in these families are more likely to be
due to incomplete genome assemblies.

@@

The Vault RNA forms a major component of the vault ribonucleoprotein
complex, this is one of the largest particles found in the vertebrate
cell; In fact, it is larger than the ribosome \cite{Kong:1999}. As yet
not much is known about the function of Vault. The Vault RNA has been
shown to be broadly conserved in metazoans
\cite{Stadler:2009}. However, in the bird lineages

@@

The human genome is littered with decaying, pseudogenised copies of
ncRNAs. The Dfam database of human repeats estimates that
approximately $54\%$ of the human genome is composed of repeat
elements \cite{Wheeler:2013}. Of these annotations XXX\% are derived
from ncRNAs, including the large Alu family of repeats, which is
derived from a retransposed SRP RNA pseudogene \cite{}.

@@

\begin{figure}[ht]
  \includegraphics[height=0.95\textheight]{figures/miRNA.pdf}
  \includegraphics[height=0.95\textheight]{figures/snoRNA.pdf}
  \caption[]{.}\label{fig:3}
\end{figure}

\begin{figure}[ht]
  \includegraphics[width=0.45\textwidth]{figures/lncRNA.pdf}
  \includegraphics[width=0.45\textwidth]{figures/tRNA.pdf}
  \caption[]{.}\label{fig:5}
\end{figure}

>>>>

%% Talking points:
%% 1.     Although the majority of classical eukaryotic ncRNAs are conserved in birds, several notable classical ncRNAs are not. Homology searches in associated genes indicate these apparent "losses" are most likely due to divergence.
%% 2.     Of the 59 snoRNA conserved between humans and yeast, we find 45 are conserved in birds. Homology searches in associated genes for the remaining non-conserved snoRNAs indicate these apparent losses are indeed due to multiple losses in several bird lineages.
%% 3.     We find that 3 families of broadly conserved vertebrate microRNA families are lost in the avian lineage.  We also annotate a number of bird-specific microRNAs.
%% 4.     Several well-characterised lncRNA families of known function are conserved between mammals and birds.
%% 5.     We discuss the conservation of HOX associated lncRNAs and several lncRNAs that have been implicated in processes associated with cancer.

%% %% Draft Abstract
%% We present the results of a large-scale bioinformatic annotation of non-coding RNAs in 48 avian genomes. We use probabilistic models of hand-curated families from the Rfam database to infer conserved RNA families within each avian genome. We supplement these annotations with predictions from the tRNA annotation tool, tRNAscan-SE, microRNAs from miRBase and selected sequences from analyses of the chicken transcriptome. Although we show extensive conservation of classical ncRNAs (e.g., tRNAs and rRNAs) and more recently discovered ncRNAs (e.g., snoRNAs and miRNAs) in birds, we also demonstrate apparent ``losses'' in several RNA families. These include the divergence of some classical ncRNAs and the genuine loss of several snoRNAs and microRNAs. In contrast, we show a signicant number of lncRNAs are surprisingly well conserved between birds and mammals including several intriguing cases where the reported mammalian lncRNA function is not conserved in birds. These combined results illustrate the utility of applying homology based methods for annotating vertebrate genomes and illustrate many complex evolutionary patterns within the avian ncRNA cohort.

@@

@@

The 5.8 S component of the ribosome in the turtle, turkey bustard,
hoatzin, flamingo, tropicbird, seriema, owl, cuckoo roller, trogon,
bee eater and falcon appears to have been lost (See
Figure~\ref{fig:1}). The rRNA repeats are frequently not assembled,
consequently it is not surprising to see ``losses'' in these \cite{Floutsakou:2013}.
Furthermore, the genomes for these species are also
low-coverage.

@@

The Vault RNA forms a major component of the vault ribonucleoprotein
complex, this is one of the largest particles found in the vertebrate
cell; In fact, it is larger than the ribosome \cite{Kong:1999}. As yet
not much is known about the function of Vault. The Vault RNA has been
shown to be broadly conserved in metazoans
\cite{Stadler:2009}. However, in the bird lineages it appears to have
either been lost or diversified.

@@

The human genome is littered with pseudogenised copies of ncRNAs. The
Dfam database of human repeats estimates that approximately $54\%$ of
the human genome is composed of repeat elements
\cite{Wheeler:2013}. Of these annotations XXX\% are derived from
ncRNAs, including the large Alu family of repeats, which is derived
from a retransposed SRP RNA pseudogene \cite{}.

@@

\begin{figure}[ht]
  \includegraphics[height=0.95\textheight]{figures/miRNA.pdf}
  \includegraphics[height=0.95\textheight]{figures/snoRNA.pdf}
  \caption[]{Heatmaps showing the prescence/abscence and approximate
    copy number of {\bf miRNA} families on the right and {\bf snoRNA} families on
    the left.}\label{fig:3}
\end{figure}

\begin{figure}[ht]
  \includegraphics[width=0.45\textwidth]{figures/lncRNA.pdf}
  \includegraphics[width=0.45\textwidth]{figures/tRNA.pdf}
  \caption[]{Heatmaps showing the prescence/abscence and approximate
    copy number of {\bf lncRNA} families on the right and {\bf tRNA}
    families on the left.}\label{fig:5}
\end{figure}
